### 2008-05-16 trunk/Papers/dl-gre-08/bisim.tex
good except for one extra line

Formulas of both $\alc$ and $\el$ are interpreted in ordinary
relational first-order models $\gM = (\Delta,\interp{\cdot})$ where
$\Delta$ is a non-empy set and $\interp{\cdot}$ is an
interpretation function such that:
$$
\begin{array}{ccl}
\interp{p} & \subseteq & \Delta  \mbox{ for $p \in \prop$}\\
\interp{R} & \subseteq & \Delta \times \Delta  \mbox{ for $R \in \rel$}\\
\interp{\neg \varphi} & = & \Delta - \interp{\varphi}\\
\interp{\varphi \sqcap \varphi'} & = & \interp{\varphi} \cap \interp{\varphi'}\\
\interp{\exists R.\varphi} & = & \{i \mid \mbox{for some } i', (i,i') \in \interp{R}\\
& & \mbox{ and } i' \in \interp{\varphi} \}.\\
\end{array}
$$

@@

Notice that similarity is not necessarily a symmetrical relation: For
instance, $f_1$ is \el-similar to $f_2$ in
Fig.~\ref{fig:dale-haddock}b, but $f_2$ is not \el-similar to $f_1$
(it satisfies the formula $\exists \mathsf{in}.\mathsf{hat}$ and $f_1$
doesn't).  However, \alc-similarity is a symmetrical relation because
the language contains negation; and indeed, $f_1$ is not \alc-similar
to $f_2$ either because it satisfied $\neg \exists
\mathsf{in}.\mathsf{hat}$.  Because \alc\ is more expressive than \el,
it is generally possible for some individual $a$ to be \el-similar but
not \alc-similar to some individual $b$, but not vice versa.

>>>>

Formulas of both $\alc$ and $\el$ are interpreted in ordinary
relational first-order models $\gM = (\Delta,\interp{\cdot})$ where
$\Delta$ is a non-empty set and $\interp{\cdot}$ is an
interpretation function such that:
$$
\begin{array}{ccl}
\interp{p} & \subseteq & \Delta  \mbox{ for $p \in \prop$}\\
\interp{R} & \subseteq & \Delta \times \Delta  \mbox{ for $R \in \rel$}\\
\interp{\neg \varphi} & = & \Delta - \interp{\varphi}\\
\interp{\varphi \sqcap \varphi'} & = & \interp{\varphi} \cap \interp{\varphi'}\\
\interp{\exists R.\varphi} & = & \{i \mid \mbox{for some } i', (i,i') \in \interp{R}\\
& & \mbox{ and } i' \in \interp{\varphi} \}.\\
\end{array}
$$

@@

Notice that similarity is not necessarily a symmetrical relation: For
instance, $f_1$ is \el-similar to $f_2$ in
Fig.~\ref{fig:dale-haddock}b, but $f_2$ is not \el-similar to $f_1$
(it satisfies the formula $\exists \mathsf{in}.\mathsf{hat}$ and $f_1$
doesn't).  However, \alc-similarity is a symmetrical relation because
the language contains negation; and indeed, $f_1$ is not \alc-similar
to $f_2$ either because it satisfies $\neg \exists
\mathsf{in}.\mathsf{hat}$.  Because \alc\ is more expressive than \el,
it is  possible for some individual $a$ to be \el-similar but
not \alc-similar to some individual $b$, but not vice versa.

### 2008-05-16 trunk/Papers/dl-gre-08/conclusion.tex
autoren rein
good except for one extra line

In this paper, we have explored the idea of viewing the generation of
singular REs as the problem of computing a DL formula with a given
extension.  We have shown how such formulas can be computed
efficiently (for \alc\ and \el) by adapting existing algorithms from
the literature.  The \el\ algorithm is able to generate 95\% of the
non-relational and 67\% of the relational REs from
\newcite{viethen06:_algor_for_gener_refer_expres}.  Both algorithms
are extremely efficient (350 ms and 140 ms respectively to generate
relational REs for all individuals in a random model with 100
individuals); to our knowledge, these are by far the fastest runtimes
for relational GRE reported in the literature.  We will make our
implementation available online.

Because they compute referring expressions for all individuals in the
domain at once, our algorithms will perform especially strongly in
static settings, such as the generation of descriptions for museum
exhibits, in which the individuals and their properties don't change
much.  However, even in more dynamic settings, our algorithms have a
chance to outperform search algorithms like Dale and Haddock's in the
average case because they can't get stuck in unproductive branches of
the search space. Nevertheless, one interesting question for future
research is how to incrementally update simulation classes when the
model changes. Similarly, it would be interesting to explore how
linguistic constraints and different attribute orderings can be taken
into account without slowing the algorithms down, and how the
algorithms can be adapted to compute REs for sets. In exploring these
extensions we will be able to draw on a rich body of literature that
considers many variants of simulation algorithms.

%Nevertheless, one interesting question for future
%research is how to adapt simulation algorithms that can
%incrementally update the simulation classes when the model changes.
%Similarly, it would be interesting to explore how linguistic
%constraints and different attribute orderings can be taken into
%account without slowing the algorithms down, and how the algorithms
%can be adapted to compute REs for sets.

In experimenting with the Viethen and Dale data, we found that
there is no single ordering that covers all human-produced
descriptions, which seems to be in contrast to Dale and Reiter's
\shortcite{Dale1995} assumption that there is only one ordering for
each given domain.  In fact, it is not even the case that each speaker
consistently uses just one ordering.  An interesting open research
question is thus what factors determine which ordering is used.
Unfortunately, both in the Viethen and Dale dataset and in the TUNA
corpus \cite{deemter06:_build_seman_trans_corpus_for}, only a minority
of referring expressions is relational, maybe because these domains
lend themselves very well to row/column style propositional REs.  We
are currently preparing an experiment to collect REs in a domain in
which relational REs will be more frequent.

>>>>

In this paper, we have explored the idea of viewing the generation of
singular REs as the problem of computing a DL formula with a given
extension.  We have shown how such formulas can be computed
efficiently (for \alc\ and \el) by adapting existing algorithms from
the literature.  The \el\ algorithm is able to generate 95\% of the
non-relational and 67\% of the relational REs from
\newcite{viethen06:_algor_for_gener_refer_expres}.  Both algorithms
are extremely efficient (350 ms and 140 ms respectively to generate
relational REs for all individuals in a random model with 100
individuals); to our knowledge, these are by far the fastest runtimes
for relational GRE reported in the literature.  We have made our
implementation available online at
\url{http://code.google.com/p/crisp-nlg/wiki/DlGre}.

Because they compute referring expressions for all individuals in the
domain at once, our algorithms will perform especially strongly in
static settings, such as the generation of descriptions for museum
exhibits, in which the individuals and their properties don't change
much.  However, even in more dynamic settings, our algorithms have a
chance to outperform search algorithms like Dale and Haddock's in the
average case because they can't get stuck in unproductive branches of
the search space. Nevertheless, one interesting question for future
research is how to incrementally update simulation classes when the
model changes. Similarly, it would be interesting to explore how
linguistic constraints and different attribute orderings can be taken
into account without slowing the algorithms down, how our algorithms
could be integrated with more standard DL T-Box inferences, and how the
algorithms can be adapted to use inverse relations or to compute REs
for sets. In exploring these extensions we will be able to draw on a
rich body of literature that considers many variants of simulation
algorithms.

In experimenting with the Viethen and Dale data, we found that there
is no single ordering that covers all human-produced descriptions,
which seems to be in contrast to Dale and Reiter's
\shortcite{Dale1995} assumption that there is only one ordering for
each given domain.  In fact, it is not even the case that each speaker
consistently uses just one ordering.  An interesting open research
question is thus what factors determine which ordering is used.
Unfortunately, both in the Viethen and Dale dataset and in the TUNA
corpus \cite{deemter06:_build_seman_trans_corpus_for}, only a minority
of referring expressions is relational, maybe because these domains
lend themselves very well to row/column style propositional REs.  We
are currently collecting REs in a domain in which propositional REs
are less preferred.

\begin{small}
\paragraph{Acknowledgments.} We are grateful to Hector Geffner (who
independently suggested to view GRE as computation of DL formulas),
Kees van Deemter, and Emiel Krahmer for interesting discussions.  We
also thank Jette Viethen and Robert Dale for making their corpus
available, and the reviewers for their comments.
\end{small}

### 2008-05-16 trunk/Papers/dl-gre-08/discussion.tex
good except for one extra line

Both the \el\ and the \alc\ algorithms took about 15 milliseconds to
compute distinguishing formulas for all 16 individuals in the Viethen
and Dale dataset.\footnote{Runtimes are meassured on a MacBook Pro
  (Intel Core 2 Duo, 2.16 GHz) running Java 1.6 beta, and we allowed the
  Java VM to warm up, i.e., just-in-time compile all bytecode, before
  taking the measurements.}

@@

Another encouraging aspect of both runtime graphs is that they seem to
be approximated well by quadratic functions.  So although the
worst-case runtime of the \el\ algorithm might be exponential, both
algorithms empirically exhibit (low) polynomial runtime on the inputs
we considered.

@@

Our GRE algorithms do not guarantee that the formula they compute can
actually be realized in language.  For example, none of the formulas
our algorithms computed in the Viethen and Dale domain contained an
atom that would commonly be realized as a noun; the property
$\mathsf{drawer}$ is never used because it applies to all individuals
in the domain.  This particular problem could easily be worked around
in a post-processing step.  However, another problem arises from the
massive use of negation in the \alc\ algorithm; it will be hard for
any realizer to find a reasonable way of expressing a formula like
$\neg \exists R.(\neg P \sqcap \neg Q)$ as a smooth noun phrase.
Although we agree with \newcite{deemter01:_gener_refer_expres} and
others that the careful use of negation and disjunction can improve
REs, these connectives must not be overused.  Thus we consider the
formulas computed by the \el\ algorithm ``safer'' with respect to
realization.

>>>>

Both the \el\ and the \alc\ algorithms took about 15 milliseconds to
compute distinguishing formulas for all 16 individuals in the Viethen
and Dale dataset.\footnote{Runtimes were measured on a MacBook Pro
  (Intel Core 2 Duo, 2.16 GHz) running Java 1.6 beta.  We allowed the
  Java VM to warm up, i.e., just-in-time compile all bytecode, before
  taking the measurements.}

@@

% Another encouraging aspect of both runtime graphs is that they seem to
% be approximated well by quadratic functions.  So although the
% worst-case runtime of the \el\ algorithm might be exponential, both
% algorithms empirically exhibit (low) polynomial runtime on the inputs
% we considered.

@@

Our GRE algorithms do not guarantee that the formula they compute can
actually be realized in language.  For example, none of the formulas
our algorithms computed in the Viethen and Dale domain contained an
atom that would commonly be realized as a noun; the property
$\mathsf{drawer}$ is never used because it applies to all individuals
in the domain.  This particular problem could easily be worked around
in a post-processing step.  However, another problem arises from the
massive use of negation in the \alc\ algorithm; it will be hard for
any realizer to find a reasonable way of expressing a formula like
$\neg \exists R.(\neg P \sqcap \neg Q)$ as a smooth noun phrase.
Although we agree with \newcite{deemter02:_gener_refer_expres} and
others that the careful use of negation and disjunction can improve
REs, these connectives must not be overused.  Thus we consider the
formulas computed by the \el\ algorithm ``safer'' with respect to
realization.

### 2008-05-16 trunk/Papers/dl-gre-08/gre.tex
good except for one extra line

Now we apply description logic to GRE.  The core claim of this paper
is that it is natural and useful to view the GRE problem as the
problem of computing a formula of description logic whose extension is
a given target set of individuals.

@@

For the rest of this paper, we assume that we are generating a
singular RE, i.e.\ the target referent $A$ will be a singleton set.
In this case, we will only be able to generate a formula that denotes
exactly $A = \{a\}$ (i.e., a RE that uniquely refers to $a$) if there
is no other individual $b$ to which $a$ is similar; otherwise, any
formula that is satisfied by $a$ is also satisfied by $b$.
Conversely, if we know that $a$ is not similar to any other
individual, then there is a formula that is satisfied by $a$ and not
by anything else; this formula can serve as a unique singular RE.  In
other words, we can reduce the $\gL$-GRE problem for a given model to
the problem of computing the $\gL$-similarity sets of this model.

@@

The pseudocode for our \alc\ algorithm is shown as
Algorithm~\ref{algo:bisim-l} and Algorithm~\ref{algo:bisim-add-alc}.
Given a model $\gM=(\Delta, \interp{\cdot})$, the algorithm computes a
set $\RE$ of $\alc$ formulas such that $\{\interp{\varphi} \mid
\varphi \in \RE\}$ is the set of \alc-similarity sets of $\gM$.  The
algorithm starts with $\RE = \{\top\}$ (where $\interp{\top} =
\Delta$), and successively refines $\RE$ by making its elements denote
smaller and smaller sets.  It maintains the invariant that at the
start and end of every iteration, $\{\interp{\varphi} \mid \varphi \in
\RE\}$ is always a partition of $\Delta$.  The algorithm iterates over
all propositional and relational symbols in $\prop$ and $\rel$ to
construct new formulas until either all formulas in $\RE$ denote
singletons (i.e., there is only one individual that satisfies them),
or no progress has been made in the previous iteration.  In each
iteration, it calls the procedure add$_\alc$($\varphi$, $\RE$), which
intersects $\varphi$ with any formula $\psi \in \RE$ which does not
denote a singleton and which is not equivalent to $\varphi$ and to
$\neg \varphi$. In this case, it replaces $\psi$ in $\RE$ by $\psi
\sqcap \varphi$ and $\psi \sqcap \neg \varphi$.

@@

Because it maintains a weaker invariant, the set $\RE$ may contain
more formulas at the same time in the \el\ algorithm than in the \alc\
algorithm.  Prima facie, this means that it has worst-case exponential
runtime, given that the whole domain has an exponential number of
subsets.  However, it is possible that a more careful analysis of the
combinatorics of the considered subsets will reveal that the algorithm
is actually polynomial.

>>>>

Now we apply description logic to GRE.  The core claim of this paper
is that it is natural and useful to view the GRE problem as the
problem of computing a formula of some description logic $\gL$ whose
extension is a given target set $A$ of individuals.

@@

For the rest of this paper, we assume that we are generating a
singular RE, i.e.\ the target set $A$ will be singleton.  In this
case, we will only be able to generate a formula that denotes exactly
$A = \{a\}$ (i.e., a RE that uniquely refers to $a$) if there is no
other individual $b$ to which $a$ is similar; otherwise, any formula
that is satisfied by $a$ is also satisfied by $b$.  Conversely, if we
know that $a$ is not similar to any other individual, then there is a
formula that is satisfied by $a$ and not by anything else; this
formula can serve as a unique singular RE.  In other words, we can
reduce the $\gL$-GRE problem for a given model to the problem of
computing the $\gL$-similarity sets of this model.  Notice that this
use of similarity sets can be seen as a generalization of van
Deemter's \shortcite{deemter02:_gener_refer_expres} ``satellite sets''
to relational descriptions.

@@

The pseudocode for our \alc\ algorithm is shown as
Algorithm~\ref{algo:bisim-l} (with $\gL=\alc$) and
Algorithm~\ref{algo:bisim-add-alc}.  Given a model $\gM=(\Delta,
\interp{\cdot})$, the algorithm computes a set $\RE$ of $\alc$
formulas such that $\{\interp{\varphi} \mid \varphi \in \RE\}$ is the
set of \alc-similarity sets of $\gM$.  The algorithm starts with $\RE
= \{\top\}$ (where $\interp{\top} = \Delta$), and successively refines
$\RE$ by making its elements denote smaller and smaller sets.  It
maintains the invariant that at the start and end of every iteration,
$\{\interp{\varphi} \mid \varphi \in \RE\}$ is always a partition of
$\Delta$.  The algorithm iterates over all propositional and
relational symbols in $\prop$ and $\rel$ to construct new formulas
until either all formulas in $\RE$ denote singletons (i.e., there is
only one individual that satisfies them), or no progress has been made
in the previous iteration.  In each iteration, it calls the procedure
add$_\alc$($\varphi$, $\RE$), which intersects $\varphi$ with any
formula $\psi \in \RE$ which does not denote a singleton and which is
not equivalent to $\varphi$ and to $\neg \varphi$. In this case, it
replaces $\psi$ in $\RE$ by $\psi \sqcap \varphi$ and $\psi \sqcap
\neg \varphi$.

@@

Because it maintains a weaker invariant, the set $\RE$ may contain
more formulas at the same time in the \el\ algorithm than in the \alc\
algorithm.  Given that $\Delta$ has an exponential number of subsets,
there is a risk that the \el\ algorithm might have worst-case
exponential runtime (although we are not aware of such worst-case
examples).  We leave a more careful runtime analysis for future work.

### 2008-05-16 trunk/Papers/dl-gre-08/introduction.tex
good except for one extra line

The generation of referring expressions (GRE) is one of the most
active and successful research areas in natural language generation.
Building upon Dale and Reiter's work \cite{dale89cooking,Dale1995},
various researchers have added extensions such as reference to sets
\cite{Stone2000}, more expressive logical connectives
\cite{deemter01:_gener_refer_expres}, and relational expressions
\cite{dale91:_gener_refer_expres_invol_relat}.

Referring expressions (REs) involving relations, in particular, have
received increasing attention recently, particularly in the context of
spatial referring expressions in situated generation (e.g.\
\cite{kelleher06:_increm_gener_of_spatial_refer}), where it seems
particularly natural to use expressions such as ``the book on the
table''.  However, the classical algorithm by
\newcite{dale91:_gener_refer_expres_invol_relat} was recently shown to
be unable to generate satisfying REs in practice
\cite{viethen06:_algor_for_gener_refer_expres}.  Furthermore, the Dale
and Haddock algorithm and most of its successors
\cite{Krahmer2003,kelleher06:_increm_gener_of_spatial_refer} are
vulnerable to the problem of ``infinite regress'', where the algorithm
jumps back and forth between generating descriptions for two related
individuals infinitely, as in ``the book on the table which supports a
book on the table \ldots''.

In this paper, we propose to view GRE as the problem of computing a
formula of description logic (DL) that denotes exactly the set of
individuals that we want to refer to.  This very natural idea has been
mentioned in passing before
\cite{Krahmer2003,gardent07:_gener_bridg_defin_descr}; however, we
take it one step further by proposing DL as an interlingua for
comparing the REs produced by different approaches to GRE.  In this
way, we can organize existing GRE approaches in an expressiveness
hierarchy.  For instance, the classical Dale and Reiter algorithms
compute purely conjunctive formulas;
\newcite{deemter01:_gener_refer_expres} extends this language by
adding the other propositional connectives, whereas
\newcite{dale91:_gener_refer_expres_invol_relat} extends it by
allowing existential quantification.

>>>>

The generation of referring expressions (GRE) is one of the most
active and successful research areas in natural language generation.
Building upon Dale and Reiter's work \cite{dale89cooking,Dale1995},
various researchers have added extensions such as reference to sets
\cite{Stone2000}, more expressive logical connectives
\cite{deemter02:_gener_refer_expres}, and relational expressions
\cite{dale91:_gener_refer_expres_invol_relat}.

Referring expressions (REs) involving relations, in particular, have
received increasing attention recently, particularly in the context of
spatial referring expressions in situated generation (e.g.\
\cite{kelleher06:_increm_gener_of_spatial_refer}), where it seems
particularly natural to use expressions such as ``the book on the
table''.  However, the classical algorithm by
\newcite{dale91:_gener_refer_expres_invol_relat} was recently shown to
be unable to generate satisfying REs in practice
\cite{viethen06:_algor_for_gener_refer_expres}.  Furthermore, the Dale
and Haddock algorithm and most of its successors (such as
\cite{kelleher06:_increm_gener_of_spatial_refer}) are vulnerable to
the problem of ``infinite regress'', where the algorithm jumps back
and forth between generating descriptions for two related individuals
infinitely, as in ``the book on the table which supports a book on the
table \ldots''.

In this paper, we propose to view GRE as the problem of computing a
formula of description logic (DL) that denotes exactly the set of
individuals that we want to refer to.  This very natural idea has been
mentioned in passing before
\cite{Krahmer2003,gardent07:_gener_bridg_defin_descr}; however, we
take it one step further by proposing DL as an interlingua for
comparing the REs produced by different approaches to GRE.  In this
way, we can organize existing GRE approaches in an expressiveness
hierarchy.  For instance, the classical Dale and Reiter algorithms
compute purely conjunctive formulas;
\newcite{deemter02:_gener_refer_expres} extends this language by
adding the other propositional connectives, whereas
\newcite{dale91:_gener_refer_expres_invol_relat} extends it by
allowing existential quantification.

### 2008-05-16 trunk/Papers/dl-gre-08/related.tex
moved figures around
good except for one extra line

\begin{figure}
  \centering
  \begin{small}
  \begin{tabular}{l|p{0.2\textwidth}}
    GRE algorithm & DL variant \\ \hline
    \newcite{Dale1995} & $\mathcal{CL}$ \\
    \newcite{deemter01:_gener_refer_expres} & $\mathcal{PL}$ \\
    \newcite{dale91:_gener_refer_expres_invol_relat} & \el \\
    \newcite{kelleher06:_increm_gener_of_spatial_refer} & \el \\
    \newcite{gardent02:_gener_minim_defin_descr} & $\mathcal{ELU}_{(\neg)}$\\
%    \newcite{Krahmer2003} & \el\ + equality between individuals
  \end{tabular}
  \end{small}
  \caption{DL variants used by different GRE algorithms.}
  \label{fig:related}
\end{figure}

Viewing GRE as a problem of generating DL formulas offers a unified
perspective of the GRE problem: It is the problem of computing a DL
formula with a given extension.  Many existing approaches can be
subsumed under this view; we have summarized this for some of them in
Fig.~\ref{fig:related}, along with the DL fragment they use.  We
already discussed some of these approaches in Section~\ref{sec:gre}.
Furthermore, the non-relational but negative and disjunctive
descriptions generated by \newcite{deemter01:_gener_refer_expres} are
simply formulas of $\mathcal{PL}$; and
\newcite{gardent02:_gener_minim_defin_descr} generalizes this into
generating formulas of $\mathcal{ELU}_{(\neg)}$, i.e., \el\ plus
disjunction and atomic negation.  The approach presented here fits
well into this landscape, and it completes the picture by showing how
to generate REs in \alc, which combines all connectives used in any of
these previous approaches.

Where our approach breaks new ground is in the way these formulas are
computed: It successively refines a decomposition of the domain into
subsets.  In this way, it is reminiscent of the Incremental Algorithm,
which in fact can be seen as the special case of the \el\ algorithm
for the non-relational case.  However, unlike
\newcite{dale91:_gener_refer_expres_invol_relat} and its successors,
such as \newcite{kelleher06:_increm_gener_of_spatial_refer},
we do not have to take special precautions to avoid infinite
regress. While Dale and Haddock's algorithm attempts to generate a RE
for a single individual for successive individuals in the model, our
algorithms consider all individuals in parallel and are guaranteed to
always terminate.

Perhaps closest in spirit to our approach is Krahmer et al.'s graph
algorithm~\shortcite{Krahmer2003}, which also computes REs by
extending them successively.  However, their subgraphs go beyond the
expressive power of \alc\ in that they can distinguish between ``the
dog that bites a dog'' and ``the dog that bites itself''.  The price
they pay for this increase in expressive power is an NP-complete
worst-case complexity.  Interestingly, Krahmer et al.\ themselves
discuss the possibility of seeing their subgraphs as formulas of
hybrid logic which are satisfied at the points where the subgraph can
be embedded; and hybrid logics can be seen as very expressive
description logics~\cite{arec:hybr05b} as well.

>>>>

Viewing GRE as a problem of generating DL formulas offers a unified
perspective of the GRE problem: It is the problem of computing a DL
formula with a given extension.  Many existing approaches can be
subsumed under this view; we have summarized this for some of them in
Fig.~\ref{fig:related}, along with the DL fragment they use.  We
already discussed some of these approaches in Section~\ref{sec:gre}.
Furthermore, the non-relational but negative and disjunctive
descriptions generated by \newcite{deemter02:_gener_refer_expres} are
simply formulas of $\mathcal{PL}$; and
\newcite{gardent02:_gener_minim_defin_descr} generalizes this into
generating formulas of $\mathcal{ELU}_{(\neg)}$, i.e., \el\ plus
disjunction and atomic negation.  The approach presented here fits
well into this landscape, and it completes the picture by showing how
to generate REs in \alc, which combines all connectives used in any of
these previous approaches.

Where our approach breaks new ground is in the way these formulas are
computed: It successively refines a decomposition of the domain into
subsets.  In this way, it is reminiscent of the Incremental Algorithm,
which in fact can be seen as the special case of the \el\ algorithm
for the non-relational case.  However, unlike
\newcite{dale91:_gener_refer_expres_invol_relat} and its successors,
such as \newcite{kelleher06:_increm_gener_of_spatial_refer},
we do not have to take special precautions to avoid infinite
regress. While Dale and Haddock's algorithm attempts to generate a RE
for a single individual for successive individuals in the model, our
algorithms consider all individuals in parallel.  It monotonically
refines a partition of the model and never needs to backtrack, and
therefore is always guaranteed to terminate.

\begin{figure}
  \centering
  \begin{small}
  \begin{tabular}{l|p{0.2\textwidth}}
    GRE algorithm & DL variant \\ \hline
    \newcite{Dale1995} & $\mathcal{CL}$ \\
    \newcite{deemter02:_gener_refer_expres} & $\mathcal{PL}$ \\
    \newcite{dale91:_gener_refer_expres_invol_relat} & \el \\
    \newcite{kelleher06:_increm_gener_of_spatial_refer} & \el \\
    \newcite{gardent02:_gener_minim_defin_descr} & $\mathcal{ELU}_{(\neg)}$\\
%    \newcite{Krahmer2003} & \el\ + equality between individuals
  \end{tabular}
  \end{small}
  \caption{DL variants used by different GRE algorithms.}
  \label{fig:related}
\end{figure}

Perhaps closest in spirit to our approach is Krahmer et al.'s graph
algorithm~\shortcite{Krahmer2003}, which also computes REs by
extending them successively.  However, their subgraphs go beyond the
expressive power of \alc\ in that they can distinguish between ``the
dog that bites a dog'' and ``the dog that bites itself''.  The price
they pay for this increase in expressive power is an NP-complete
worst-case complexity.  Interestingly, Krahmer et al.\ themselves
discuss the possibility of seeing their subgraphs as formulas of
hybrid logic which are satisfied at the points where the subgraph can
be embedded; and hybrid logics can be seen as very expressive
description logics as well \cite{arec:hybr05b}.
